Formal models of biological networks in \animo2{}

### 2013-12-16 Paper.tex
Added more on model checking

\def\ta{Timed Automaton}
\def\tas{Timed Automata}

@@

\title{Formal models of biological networks in ANIMO 2} %Building formal models of biological networks avoiding formal training}

@@

In this paper we present ANIMO 2.0, an updated version of ANIMO which allows for a considerably
more efficient analysis of models. ANIMO 2.0 also introduces new features, such as the possibility to interrogate
an ANIMO model with model checking queries represented as human language sentences instead of
mathematical formulae. The case studies presented in this paper are centered around possible usage scenarios
for ANIMO, and aim at illustrating how a biologist can profit from the tool in its latest iteration.
An installation and user manual for ANIMO 2.0 is available at \url{http://fmt.cs.utwente.nl/tools/animo/manual.html}.

\subsection*{A short introduction to ANIMO}
{\Large [[briefly describe ANIMO, its modelling assumptions and its intended use. Probably good idea
to refer to Gene paper]]}

@@

\subsection*{Testing the performances of ANIMO 2.0}\label{subsec:model-checking}
% \subsection*{Reaction-centered VS reactant-centered}\label{sec:animo-comparison}
% We will now apply some basic model checking queries to the case study presented in~\cite{animo-bibe},
% measuring the performances of the two modeling approaches. This will allow us to
% evaluate the benefit brought by the shift in perspective from a reaction- to a reactant-centered model.
In addition to some new features, ANIMO 2.0 provides a boost in terms of performances
(see Methods section for details).
In order to measure how significant the performance increase is, we applied a number of
ANIMO analysis queries to a case study, comparing the performances of ANIMO 2.0 with
the previous version of the tool. Moreover, ANIMO 2.0 provides the biolgist with
a proper access to the formal technique of model checking~\cite{model-checking}. As we will show, the
performance improvement is even more significant in this case, making this type of formal
analysis more accessible.

@@

\subsubsection*{Simulation-based analysis}
The most basic analysis that can be performed with ANIMO is based on simulating the behaviour
of the network during a given period of time (see also the analyses
of the Drosphila and TNF-$\alpha$/EGF models above). As this type of analysis is also normally
used when looking for a match between model parameters and experimental data, the computation
speed of a simulation run deeply affects the level of interactivity perceived by the user.
In order to make
the modeling approach in ANIMO as smooth as possible, it is desirable to decrease dead times,
i.e. significantly reduce simulation costs.
% UPPAAL is used in ANIMO to obtain simulation runs from a \tas\ model. While that process is
% very fast, 
% In this experiment, we query UPPAAL for simulation runs on the models
Our first performance test compares the execution time for simulation runs on the \tas\ models
generated by ANIMO applying the old (i.e., ANIMO 1.0) and new (ANIMO 2.0) modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the initial activity level of nodes NGF to 15/15, and EGF to 0/15. We have chosen this configuration as the
model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
This makes the model more demanding when generating the state space
because having more reactions occur means having more states in the transition system underlying the \tas\ model.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $92 \%$ less time with the new version of ANIMO, using $87 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network. This advantage is particularly noticeable when analysing
more complex networks.
{\Large [[Add also lines for drosophila and TNFa-EGF models, that are larger and make the performance increment look even more impressive]]}

\subsubsection*{A good reason to use formal methods: model checking}
% Next, we set out to test the model checking performances using the old \tas\ model used in ANIMO and
% the new one introduced in ANIMO 2.
To show why one should use a formal model like \tas, we demonstrate the application
one of the biggest advantages of \tas: model checking~\cite{model-checking}.

The technique of model checking consists in verifying the truth of some properties
that describe the behaviour of a model. This is usually done by exploring every
possible execution path of the model starting from its initial configuration. The set of all
the configurations that can be reached by a model during all its possible executions
is called \emph{state space}. The properties on the evolution of the model are
usually expressed in a formal language such as computation tree logic (CTL,~\cite{ctl}),
which allows to formally describe how and when a property should hold. Examples of such
properties can be expressed in natural language as: ``the activity of node A
stays always above 20'', ``it is possible to reach a state where the activity of both nodes A and B
is at least 60'', ``whenever the activity of node A surpasses 20, the activity of node
B will eventually reach at least 40''.

Model checking performances in ANIMO 2.0 have been tested using the case study of EGF-NGF signalling as reference.
We compared the execution times and memory requirements in ANIMO and ANIMO 2.0 for a number of interesting queries:
\begin{itemize}
  \item (1) and (2): {\tt A[] not deadlock}. The model continues to execute indefinitely ({\tt A} refers to all
      possible paths in the transition system of the model, and {\tt []} asks the property to always hold along a path).
  \item (3): {\tt RKIP $<$ 10 --$>$ ERK $>=$ 40}. After RKIP (Raf kinase inhibitory protein) activity has been lowered, ERK activity increases. As in the model RKIP
      has 20 levels of granularity and ERK has 100 levels, {\tt RKIP $<$ 10} means that RKIP is less than half active, and
      {\tt ERK $>=$ 40} means that ERK activity is at least $40 \%$.
  \item (4): {\tt E<> RKIP $<$ 10}. Find a point when RKIP is low ({\tt E} asks for the existence of at least one path
      for which the property holds, while {\tt <>} requires the property to hold at least once along a given path).
      This query is expected to generate a trace, the last point of which will be used as initial configuration for model checking queries (5) and (6).
  \item (5): {\tt A[] ERK $<$ 70} and (6): {\tt A[] ERK $>$ 35}. Once RKIP activity has significantly decreased, ERK activity is sustained at an intermediate level.
\end{itemize}
The initial conditions are:
\begin{itemize}
  \item (1): EGF = 15/15 and NGF = 0/15, all others as the original configuration, corresponding to the treatment condition with
	    100 ng/ml EGF.
  \item (2) - (4): EGF = 0/15, NGF = 15/15, all others as the original configuration, corresponding to the treatment condition with
	    50 ng/ml NGF\footnote{In the laboratory experimental setting, NGF is used at a lower concentration than EGF, but it is still enough to saturate all NGF receptors, which
	    are rarer than EGF receptors.}.
  \item (5) and (6): all activities as in the last state of the trace computed from query~(4).
\end{itemize}

@@

Moreover, our experiments point out that ANIMO 2.0 considerably lowers the memory
requirements for the model. This is not only due to the absence of possibly large precomputed time tables,
which can contain up to ten thousand elements each.\footnote{In the model used in ANIMO 1.0, having a reaction depend on two 100-levels reactants
lead to the generation of two time tables containing $100 \times 100 = 10\,{}000$ elements each.}
% Indeed, this point was further investigated by implementing a reaction-centered model which
% avoids the use of tables and instead makes on-the-fly computations of the time bounds with the same number representation as in the reactant-centered model.
% This resulted in improved performances in the cases of reachability and simulation-based queries, with memory requirements similar to the ones for
% the reactant-centered model. However, in all other cases a much larger amount of memory (more than 2 Gb) was used
% with respect to the table-based implementation of the same model, without leading to appreciable benefits in terms of execution time:
% in some cases performances were noticeably deteriorated. These findings support the intuitive idea that a reactant-centered approach
% leads in general to fewer events in a model, and to a smaller state space.

@@

\subsection*{Model checking for bigger models}
When a model becomes particularly complex, and it represents many events occurring concurrently,
the state space of the model tends to grow exponentially in size, dramatically
increasing the time required for its analysis. To avoid this \emph{state space explosion},
other approaches to model checking are necessary. One of such approaches is based on statistical
considerations and allows to obtain approximated answers with a given confidence level in far less time
than most exact model checking techniques: it is called statistical model checking. As the newer
versions of UPPAAL include also a statistical model checking engine~\cite{uppaal-smc}, we are
able to obtain answers in case a model displays stochastic behavior. Models of this type 
usually represent non-deterministic behavior with by defining a probabilistic distribution for the
choice of alternative evolution paths. In ANIMO, it is possible to use a uniform probabilistic choice
to decide the value of an interaction speed (chosen within the user-given lower and upper bounds)
and the initial activity level of a node. If no such variability is introduced, ANIMO models are deterministic,
i.e. they will always exhibit the same behavior. As an example application of the statistical model checking
approach to an ANIMO model, we introduce variability in the initial quantities of TNF$\alpha$ and EGF
in the model of Figure~\ref{fig:large-model-complete}, letting them be chosen randomly on the interval $\{0, \dots, 100\}$.
we then interrogate the model to know what is the probability that MK2 becomes activated before 40 minutes
have passed. An answer is found whithin XY seconds, and informs us that \dots.
Obtaining an answer to the corresponding exact query (``is it possible to reach a state in which MK2 is active
before 40 minutes?'') would take too long even if using the new model introduced in ANIMO 2.0.

\subsection*{User interface for model checking in ANIMO 2}\label{sec:animo-model-checking-ui}
In order to allow a non-expert user to profit from the power of model checking, we have implemented
a template-based user interface to define queries directly inside the ANIMO Cytoscape plug-in:
Figure~\ref{fig:model-checking-ui} shows the interface for composing a model checking query in ANIMO.
The mappings between user interface templates and actual model checking queries are inspired
on the ones proposed in~\cite{hidde-templates}, and are shown in Table~\ref{tab:model-checking-templates}.

@@

\section*{Discussion}

\subsection*{Comparison between ANIMO and other modelling tools}\label{suppl:comparison-table}
Different formalisms are in use in the field of computational
modelling of biological systems, each with their specific characteristics.
Many of these formalisms have been implemented into
software tools to support modelling efforts. In order to compare
ANIMO with existing tools, we have selected a number of mathematical formalisms,
each connected to a supporting tool. With an emphasis on the modelling
process rather than the final model, we compared these tools on
the basis of the following parameters:

\begin{enumerate}
  \item {\bf Hidden formalism:} a knowledge of the underlying formalism is not required in order to use the tool
  \item {\bf Visual modelling:} the tool allows the user to model using a visual interface, and is not exclusively
      founded on formula-, text- or table-based input forms
  \item {\bf Qualitative parameters:} parameters for reactions can be input as approximated estimations, and not exclusively as numbers
  \item {\bf Tight coupling with topology:} models are tightly and clearly coupled to the networks they represent, showing the visual
      representation of the model in a shape similar or comparable to the representation currently used by biologists
      for signalling pathways
  \item {\bf User-chosen granularity:} if discretization is applied during the modelling process, the user can change the granularity
      with which such discretization is made, possibly for each component of the model separately
\end{enumerate}
Table~\ref{tab:tool-comparison} shows the comparison between ANIMO and the selected tools.

@@

The smaller amount of parameters needed to build an ANIMO model, if compared e.g. to
a differential equation-based approach, makes the modelling process faster, allowing
the biologist to obtain interesting results in less time. However, a less
parameter-intensive modelling approach is also less precise: in our case we abstract
sequences of biochemical reactions into activation or inactivation interactions,
simplifying any relation between two entities in a network as either activating or
inhibiting. While this allows for some simplifications in the networks and the underlying
formal models, it does not allow to represent in full detail more specific concepts
such as reactant concentration or reaction affinity. In case such concepts are of
primary importance, we encourage the interested reader to turn to more precise
modelling approaches, such as those based on differential equations.

All reactants modelled in ANIMO are assumed to be present at the same
(unitary) concentration. While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions. Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degratation being represented as inhibition. As stated before,
it cannot be expected from such models to be a completely realistic representation of their target biological
processes, but they can still be an useful tool. This can be seen for example in the Drosophila
model in the Results section, where mRNA and protein concentrations were abstracted to
activity-driven processes in the ANIMO model.

\section*{Conclusions}
{\Large [[Possibly further reduce these subsections]]}
% The ultimate aim of research projects is to solve a problem or get
% the answer to a scientific question. In biology, an in-depth understanding of
% the relevant biological system is an important step towards this goal. Successive 
% repetitions of the empirical cycle result in a stepwise increase in understanding,
% until the goal is reached. For complex biological systems, computational modelling is 
% indispensable in this process. The mere act of creating 
% a computational model based on prior knowledge, experimental data and hypotheses 
% assists in gaining more insight in the system. 
% 
% %Furthermore, \emph{in silico} experiments 
% %with the model can be used to make predictions that cannot be made by the human brain. 
% 
% We developed a new modelling approach by proposing a series of abstractions from the detailed 
% molecular mechanisms of biological systems. These abstractions reduce the need for kinetic 
% parameters, while preserving enough expressivity for a useful description of the dynamic 
% behaviour of biological networks. A novel modelling tool, ANIMO, allows 
% effective use of this approach and enables an intuitive construction of formal models.

ANIMO is not the first modelling tool to provide an interface to a
modelling formalism. Such interfaces exist in many other tools (as shown in Tab.~\ref{tab:tool-comparison}). With its
focus on user-friendliness and intuitive modelling, ANIMO's main contribution lies 
in making computational modelling more accessible to experts in biology.
Making use of the visual
interface provided by Cytoscape, network representations subscribe to biological conventions. 
Model parameters are kept to a minimum and can be directly accessed by mouse-clicking on 
nodes and edges. Because of the automatic translation of the network topology and 
user-defined parameters into an underlying formal model, training in the use of formal methods 
is not needed.

\subsection*{Final remarks on the models}
In the Results section, we described the construction of an ANIMO
model of the ciracadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar 
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modelling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model by~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO, instead, a number of network
nodes is drawn for the molecules involved. 
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter 
that defines the strength of each
interaction. This is a more intuitive approach to construct a model.
Further contributing to an interactive modelling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.

In the Results section we also showed the construction of an executable model
of signalling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modelling studies, based on partial least-squares regression and fuzzy 
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model 
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modelling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct in the dynamic behaviour of the network. Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behaviour than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behaviour led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behaviour of JNK1. Although this leads to a representative
description of the dynamic behaviour of JNK1, peaks in protein activity at early time points were
not reproduced by the fuzzy logic model. Moreover, it gives no insight in the molecular interactions 
that are involved in activation or inhibition.

>>>>

\def\ta{Timed Automaton}
\def\tas{Timed Automata}
\def\animo2{ANIMO 2}

@@

\title{Formal models of biological networks in \animo2{}} %Building formal models of biological networks avoiding formal training}

@@

In this paper we present \animo2{}, an updated version of ANIMO which allows for a considerably
more efficient analysis of models. \animo2{} also introduces new features, such as the possibility to interrogate
an ANIMO model with model checking queries represented as human language sentences instead of
mathematical formulae. The case studies presented in this paper are centered around possible usage scenarios
for ANIMO, and aim at illustrating how a biologist can profit from the tool in its latest iteration.
An installation and user manual for \animo2{} is available at \url{http://fmt.cs.utwente.nl/tools/animo/manual.html}.

\subsection*{A short introduction to ANIMO}
{\Large [[briefly describe ANIMO, its modelling assumptions and its intended use. Refer to Gene paper?]]}

@@

\subsection*{Testing the performances of \animo2{}}\label{subsec:model-checking}
% \subsection*{Reaction-centered VS reactant-centered}\label{sec:animo-comparison}
% We will now apply some basic model checking queries to the case study presented in~\cite{animo-bibe},
% measuring the performances of the two modeling approaches. This will allow us to
% evaluate the benefit brought by the shift in perspective from a reaction- to a reactant-centered model.
The most important development we introduce in \animo2{} is a faster analysis process,
thanks to the more efficient modelling approach (see Methods section).
In order to measure how significant the performance increase is, we applied a number of
analysis queries to a case study, comparing the performances of \animo2{} with
the previous version of the tool. As \animo2{} provides the biolgist with
a proper access to the formal technique of model checking~\cite{model-checking},
we tested the model performances also in this setting. We will show that the
performance improvement is even more significant in the case of model checking, making the use of \tas\
as formal foundation for modeling biological processes both more accessible and more profitable.

@@

\subsubsection*{Simulation-based analysis}
The simplest analysis that can be performed with ANIMO is based on simulating the behavior
of the network during a given period of time (see also the analyses
of the Drosophila and TNF-$\alpha$/EGF models above). As this type of analysis is also normally
used when looking for a match between model parameters and experimental data, the computation
speed of a simulation run deeply affects the level of interactivity perceived by the user.
In order to make
the modeling approach in ANIMO as smooth as possible, it is desirable to decrease dead times,
i.e. significantly reduce simulation costs.
% UPPAAL is used in ANIMO to obtain simulation runs from a \tas\ model. While that process is
% very fast, 
% In this experiment, we query UPPAAL for simulation runs on the models
Our first performance test compares the execution time for simulation runs on the \tas\ models
generated by ANIMO applying the old (i.e., ANIMO 1.0) and new (\animo2{}) modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the initial activity level of nodes NGF to 15/15, and EGF to 0/15. We have chosen this configuration as the
model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
This makes the model more demanding when generating the state space
because having more reactions occur means having more states in the transition system underlying the \tas\ model.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $92 \%$ less time with the new version of ANIMO, using $87 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network. This advantage is particularly noticeable when analyzing
more complex networks.
{\Large [[Add also lines for drosophila and TNFa-EGF models, that are larger and make the performance increment look even more impressive]]}

\subsubsection*{A good reason to use formal methods: model checking}
% Next, we set out to test the model checking performances using the old \tas\ model used in ANIMO and
% the new one introduced in \animo2{}.
To show why one should use a formal model like \tas, we demonstrate the application
one of the biggest advantages of \tas: model checking~\cite{model-checking}.

The technique of model checking consists in verifying the truth of some properties
that describe the behavior of a model. This is usually done by exploring every
possible execution path of the model starting from its initial configuration. The set of all
the configurations that can be reached by a model during all its possible executions
is called \emph{state space}. The properties on the evolution of the model are
usually expressed in a formal language such as computation tree logic (CTL,~\cite{ctl}),
which allows to formally describe how and when a property should hold. Examples of such
properties can be expressed in natural language as: ``the activity of node A
stays always above 20'', ``it is possible to reach a state where the activity of both nodes A and B
is at least 60'', ``whenever the activity of node A surpasses 20, the activity of node
B will eventually reach at least 40''.

Model checking performances in \animo2{} have been tested using the case study of EGF-NGF signaling as reference.
We compared the execution times and memory requirements in ANIMO and \animo2{} for a number of interesting queries:
\begin{itemize}
  \item (1) and (2): {\tt A[] not deadlock}. The model continues to execute indefinitely ({\tt A} refers to all
      possible paths in the transition system of the model, and {\tt []} asks the property to always hold along a path).
  \item (3): {\tt RKIP $<$ 10 --$>$ ERK $>=$ 40}. After RKIP (Raf kinase inhibitory protein) activity has been lowered, ERK activity increases. As in the model RKIP
      has 20 levels of granularity and ERK has 100 levels, {\tt RKIP $<$ 10} means that RKIP is less than half active, and
      {\tt ERK $>=$ 40} means that ERK activity is at least $40 \%$.
  \item (4): {\tt E<> RKIP $<$ 10}. Find a point when RKIP is low ({\tt E} asks for the existence of at least one path
      for which the property holds, while {\tt <>} requires the property to hold at least once along a given path).
      This query is expected to generate a trace, the last point of which will be used as initial configuration for model checking queries (5) and (6).
  \item (5): {\tt A[] ERK $<$ 70} and (6): {\tt A[] ERK $>$ 35}. Once RKIP activity has significantly decreased, ERK activity is sustained at an intermediate level.
\end{itemize}
The initial conditions are:
\begin{itemize}
  \item (1): EGF = 15/15 and NGF = 0/15, all others as the original configuration, corresponding to the treatment condition with
	    100 ng/ml EGF.
  \item (2) - (4): EGF = 0/15, NGF = 15/15, all others as the original configuration, corresponding to the treatment condition with
	    50 ng/ml NGF\footnote{In the laboratory experimental setting, NGF is used at a lower concentration than EGF, but it is still enough to saturate all NGF receptors, which
	    are rarer than EGF receptors.}.
  \item (5) and (6): all activities as in the last state of the trace computed from query~(4).
\end{itemize}

@@

Moreover, our experiments point out that \animo2{} considerably lowers the memory
requirements for the model. Further investigation on this point (see Methods section)
makes us believe that the main gain is indeed due to the new modeling perspective used in \animo2{}.

@@

\subsection*{Model checking for bigger models}
When a model becomes particularly complex, its state space tends to grow considerably, dramatically
increasing the model checking times. The most prominent cases of such state space growth
offer the so-called \emph{state space explosion} challenge: in such cases,
it is often necessary to find other approaches to conventional model checking. One of such approaches is based on statistical
considerations and allows to obtain approximated answers with a given confidence level in far less time
than most exact model checking techniques: it is called statistical model checking. The newer
versions of UPPAAL include also a statistical model checking engine~\cite{uppaal-smc}, enabling
us to obtain approximate answers for the bigger \tas\ models, when exploring the whole state space would cost too much.
Models that particularly benefit from statistical model checking 
usually represent non-deterministic behavior by defining probabilistic distributions for the
choice of alternative evolution paths. Such paths are potentially infinite in number, and their
exhaustive exploration would be mostly impossible using exact model checking.

Basic models built in ANIMO are deterministic, i.e. they will always exhibit the same behavior.
However, \animo2{} now allows to define the $k$-value representing the scale factor for an interaction
with a coupled \emph{uncertainty} level. For example, consider an interaction $A \rightarrow B$ described
by scenario 1 (i.e., its speed depends only on the activity level of A), with parameter $k = 0.02$. If the model
behaves as expected but we want to test it for robustness, we can ``relax'' the value of $k$ by letting
it be equally possibly inside a $\pm 5\%$ interval around the original value. Setting the uncertainty for the $k$-value
of $A \rightarrow B$ at $5 \%$ induces a uniformly distributed choice on that $k$-value in the underlying \tas\ model.
The value for $k$ is chosen as a first step at the beginning of the evolution of the model,
sampling from a uniform random distribution in the interval $[0.95 \times k, 1.05 \times k]$, and using the chosen value
along the rest of that model's execution path. Another way to add non-determinism to \animo2{} models is to
let the initial activity of one (or more) nodes range inside user-defined bounds. For example, the initial activity of $A$
could be chosen among values ranging from 20/100 to 70/100. Similarly to what is done with uncertainty in $k$-values,
the underlying \tas\ model will make a starting choice for the initial activity of $A$ based on an uniform distribution
on the interval $\{ 20, 21, \dots, 69, 70\}$. If using both the non-deterministic settings described in the example,
two simulations could use settings of $k = 0.0202$ ($k + 1\%$), $A = 25$ or $k = 0.0194$ ($k - 3\%$),
$A = 55$ with equal probability, generating different simulation courses.

While uncertainty for $k$-values can be
used to test a model's robustness against parameter variations, the possibility to add non-determinism in the choice of
initial activities can be used in \animo2{} as a basic \emph{dosage-based therapy}. Assuming that some inputs to an ANIMO model represent
drugs, we can let their initial activity (which may represent drug concentration) range over different intervals, and then interrogate ANIMO
with model checking queries to understand whether it is possible to obtain a desirable state in the network.

As an example application of the statistical model checking
approach to an ANIMO model, we introduce variability in the initial activity of TNF$\alpha$
in the model of Figure~\ref{fig:large-model-complete}, letting it be chosen randomly on the interval $\{0, \dots, 100\}$.
The initial activity of EGF is kept at 100/100.
We then interrogate the model to know what is the probability that MK2 activity is higher than 40/100 before 30 minutes
have passed. An answer is found within 9 seconds, after performing 111 simulations,
and informs us that the probability of the requested event lies in the interval $[0.87, 0.97]$ with confidence of $0.95$.
This implicitly tells us that the minimum activity of TNF$\alpha$ that still generates the useful result (MK2 $>$ 40/100
in 30 minutes) lies between $100-97 = 3$ and $100-87 = 13$ with a probability of $95\%$.
To have a more precise idea, we reduce the interval of TNF$\alpha$ to $\{0, \dots, 13\}$. After 27 seconds and 355 simulation runs,
we obtain that the probability to activate MK2 is inside the interval $[0.62, 0.72]$ with confidence of $0.95$, which gives
us a $95\%$ chance that the minimum amount of TNF$\alpha$ activity to activate MK2 is between $13 - 0.72 \times 13 = 4$
and $13 - 0.62 \times 13 = 5$. Indeed, fixing the interval for TNF$\alpha$ to $\{0, 1, 2, 3\}$ gives a probability
to activate MK2 in the interval $[0, 0.02]$ (22 seconds of computation, 263 runs, $0.99\%$ confidence), while choosing an interval of $\{4, 5, 6, 7 \}$
gives a probability in $[0.98,1]$ (22 seconds, 263 runs, $0.99\%$ confidence).
% If we lower the maximum amount of TNF$\alpha$ to 10, however, we obtain that 
% Obtaining an answer to the corresponding exact query (``is it possible to reach a state in which MK2 is active
% before 40 minutes?'') would take too long even if using the new model introduced in \animo2{}.

\subsection*{User interface for model checking in \animo2{}}\label{sec:animo-model-checking-ui}
In order to allow a non-expert user to profit from the power of model checking, we have implemented
a template-based user interface to define queries directly inside the ANIMO Cytoscape plug-in:
Figure~\ref{fig:model-checking-ui} shows the interface for composing a model checking query in ANIMO.
The mappings between user interface templates and actual model checking queries are inspired
on the ones proposed in~\cite{hidde-templates}, and are shown in Table~\ref{tab:model-checking-templates}.

@@

\section*{Discussion}
\subsection*{Final remarks on the models}
In the Results section, we described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar 
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modeling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model by~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO, instead, a number of network
nodes is drawn for the molecules involved. 
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter 
that defines the strength of each
interaction. This is a more intuitive approach to construct a model.
Further contributing to an interactive modeling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.

In the Results section we also showed the construction of an executable model
of signaling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modeling studies, based on partial least-squares regression and fuzzy 
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model 
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modeling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct in the dynamic behavior of the network. Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behavior than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behavior led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behavior of JNK1. Although this leads to a representative
description of the dynamic behavior of JNK1, peaks in protein activity at early time points were
not reproduced by the fuzzy logic model. Moreover, it gives no insight in the molecular interactions 
that are involved in activation or inhibition.

In this study, we used the same data set and performed a single round of the empirical cycle.
This cycle starts off with the experiments carried out by~\cite{pathway-compendium}. We used the
resulting experimental data, together with knowledge from curated databases~\cite{kegg,phosphosite}
to construct an executable model of the biological system.
In contrast with the two approaches described above, ANIMO is aimed at the construction of
more mechanistic models, mimicking biochemical interactions \emph{in silico}. This way of modeling
gives a different type of insight. In the process of model construction, we extended a
prior-knowledge network with time-dependent extracellular crosstalk that has been reported
previously~\cite{pathway-autocrine}. To come up with possible explanations for a disagreement
between the model and the experimental data, two additional layers of
crosstalk were introduced, at the signal transduction and transcriptional level. These modifications 
improved the fit of the model to the data and can be interpreted as novel testable hypotheses.
Finally, we proposed new experiments that could be carried out to test these hypotheses, closing the empirical cycle. 
Together, our model sheds more light on the intricate
entanglement between the TNF$\alpha$ and EGF pathways at multiple cellular levels.
But above all,  the model provides an excellent starting point for further investigation.
% Every new round in the empirical cycle will lift the understanding of the system to a higher level,
% leading to an incremental build-up of knowledge and an upward empirical spiral. Being intuitively
% accessible, ANIMO models facilitate sharing knowledge within and between groups and encourage collaborations.

\subsection*{Comparison between ANIMO and other modeling tools}\label{suppl:comparison-table}
Different formalisms are in use in the field of computational
modeling of biological systems, each with their specific characteristics.
Many of these formalisms have been implemented into
software tools to support modeling efforts. In order to compare
ANIMO with existing tools, we have selected a number of mathematical formalisms,
each connected to a supporting tool. With an emphasis on the modeling
process rather than the final model, we compared these tools on
the basis of the following parameters:

\begin{enumerate}
  \item {\bf Hidden formalism:} a knowledge of the underlying formalism is not required in order to use the tool
  \item {\bf Visual modeling:} the tool allows the user to model using a visual interface, and is not exclusively
      founded on formula-, text- or table-based input forms
  \item {\bf Qualitative parameters:} parameters for reactions can be input as approximated estimations, and not exclusively as numbers
  \item {\bf Tight coupling with topology:} models are tightly and clearly coupled to the networks they represent, showing the visual
      representation of the model in a shape similar or comparable to the representation currently used by biologists
      for signaling pathways
  \item {\bf User-chosen granularity:} if discretization is applied during the modeling process, the user can change the granularity
      with which such discretization is made, possibly for each component of the model separately
\end{enumerate}
Table~\ref{tab:tool-comparison} shows the comparison between ANIMO and the selected tools.

@@

The smaller amount of parameters needed to build an ANIMO model, if compared e.g. to
a differential equation-based approach, makes the modeling process faster, allowing
the biologist to obtain interesting results in less time. However, a less
parameter-intensive modeling approach is also less precise: in our case we abstract
sequences of biochemical reactions into activation or inactivation interactions,
simplifying any relation between two entities in a network as either activating or
inhibiting. While this allows for some simplifications in the networks and the underlying
formal models, it does not allow to represent in full detail more specific concepts
such as reactant concentration or reaction affinity. In case such concepts are of
primary importance, we encourage the interested reader to turn to more precise
modeling approaches, such as those based on differential equations.

All reactants modeled in ANIMO are assumed to be present at the same
(unitary) concentration. While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions. Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degradation being represented as inhibition. As stated before,
it cannot be expected from such models to be a completely realistic representation of their target biological
processes, but they can still be an useful tool. This can be seen for example in the Drosophila
model in the Results section, where mRNA and protein concentrations were abstracted to
activity-driven processes in the ANIMO model.

\section*{Conclusions}
% The ultimate aim of research projects is to solve a problem or get
% the answer to a scientific question. In biology, an in-depth understanding of
% the relevant biological system is an important step towards this goal. Successive 
% repetitions of the empirical cycle result in a stepwise increase in understanding,
% until the goal is reached. For complex biological systems, computational modelling is 
% indispensable in this process. The mere act of creating 
% a computational model based on prior knowledge, experimental data and hypotheses 
% assists in gaining more insight in the system. 
% 
% %Furthermore, \emph{in silico} experiments 
% %with the model can be used to make predictions that cannot be made by the human brain. 
% 
% We developed a new modelling approach by proposing a series of abstractions from the detailed 
% molecular mechanisms of biological systems. These abstractions reduce the need for kinetic 
% parameters, while preserving enough expressivity for a useful description of the dynamic 
% behaviour of biological networks. A novel modelling tool, ANIMO, allows 
% effective use of this approach and enables an intuitive construction of formal models.

ANIMO is not the first modeling tool to provide an interface to a
modeling formalism. Such interfaces exist in many other tools (as shown in Tab.~\ref{tab:tool-comparison}). With its
focus on user-friendliness and intuitive modeling, ANIMO's main contribution lies 
in making computational modeling more accessible to experts in biology.
Making use of the visual
interface provided by Cytoscape, network representations subscribe to biological conventions. 
Model parameters are kept to a minimum and can be directly accessed by mouse-clicking on 
nodes and edges. Because of the automatic translation of the network topology and 
user-defined parameters into an underlying formal model, training in the use of formal methods 
is not needed.

\subsection*{\animo2{}}
We have presented here how the new and improved \animo2{} 
provides an interactive approach to model analysis and a way of exploiting the power of model checking.
Thanks to the increased performances,
we are able to obtain answers to model checking queries in a matter of seconds.
Coherently with the user friendliness of the tool, we ensure that the biologist can access the features
of model checking without the need to directly deal with \tas\ models.
In this way, ANIMO acts as an intermediary between the biologist and a formal
representation of biological signaling pathways, letting the experts concentrate
on investigating the mechanisms of cellular responses.
